Parsing Postfix log files

### 2009-03-19 doc/logparser-implementation.tex
Dealt with more XXX items.

    \item [regex] The regex to recognise log lines with.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of captured data; and makes each regex largely
        self-documenting.  For efficiency the keywords are expanded and
        every rule's regex is compiled before attempting to parse the log
        file; otherwise every regex would be recompiled each time it was
        used, resulting in a large, data dependent slowdown, as shown in
        \sectionref{Caching compiled regexes}.

@@

XXX MAKE ALL ACTION DESCRIPTIONS FLOW BETTER\@.

@@

    \item [CLEANUP\_PROCESSING] \daemon{cleanup} processes every mail that
        passes through Postfix; details of what it does are available in
        \sectionref{Postfix Daemons}.  This action saves all data captured
        by the rule's regex if the log line has not come after a timeout
        (see \sectionref{discarding cleanup log lines}); it also creates
        the mail if necessary (see \sectionref{pickup logging after
        cleanup}).

@@

    \item [CONNECT] Handle a remote client connecting: create a new
        connection indexed by \daemon{smtpd} \acronym{pid}, and save the
        data captured by the rule's regex (\acronym{IP} address and
        hostname of the client).  If a connection already exists it is
        treated as a symptom of a bug in \parsername{}, and the action will
        issue a warning containing the full contents of the existing
        connection plus the log line that has just been parsed.

@@

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check (see
        \sectionref{out of order log lines}).  To ensure that the mail can
        be committed the EXPIRY action sets a flag marking the mail as
        expired; this flag later causes the valid combinations check to be
        skipped.  The mail will be committed when a log line triggers the
        \texttt{COMMIT} action; if the \texttt{COMMIT} action has already
        run and refused to enter the mail in the database because it failed
        the valid program combinations check, the \texttt{SAVE\_DATA}
        action will re-invoke the \texttt{COMMIT} action the next time it
        is invoked.

@@

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        \sectionref{pickup logging after cleanup}); if it does not exist it
        will be created.

    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
        it kills all \daemon{smtpd} processes, requiring any active
        connections to be cleaned up, entered in the database, and deleted
        from the state tables.  Postfix probably kills all the other
        components too, but the parser is only affected by \daemon{smtpd}
        processes exiting.

@@

    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
        signal; this is processed in the same way as the
        \texttt{SMTPD\_DIED} action.

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations; after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts that should occur in
        the intervening hours, e.g.\ timeouts for DNS requests, timeouts
        reading data from the client.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded.  The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See \sectionref{timeouts during data
        phase} for the gory details.

@@

    \item [UNINTERESTING] This rule just returns successfully; it is used
        when a line needs to be parsed for completeness but does not either
        provide any useful data or require anything to be done.

@@

XXX NEED AN INTRODUCTION\@.

@@

The theoretical aspects of this component are covered in detail in
\sectionref{rules in architecture}; this section is concerned with the
practical aspects of how rules are implemented and used in \parsername{}.
The structure of the rules table has already been documented in
\sectionref{rules table}; that abstract description will not be repeated
here, but should be fleshed out by the concrete example rule in
\sectionref{example rule in implementation}.  The process of creating new
rules from unparsed log lines is dealt with in \sectionref{creating new
rules in implementation}; the implementation of the utility supplied with
\parsername{} to create regexes from unparsed log lines is also described
in that section.  The rule conditions used in \parsername{} are explained
next, and the final portion of this section provides some
suggestions for how to detect overlapping rules.

@@

XXX WRITE THIS BIT, AND REDUCE THE ENTRY IN THE RULES TABLE\@.

    The regex to recognise log lines with.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of captured data; and makes each regex largely
        self-documenting.  For efficiency the keywords are expanded and
        every rule's regex is compiled before attempting to parse the log
        file; otherwise every regex would be recompiled each time it was
        used, resulting in a large, data dependent slowdown, as shown in
        \sectionref{Caching compiled regexes}.

>>>>

    \item [regex] The regex to recognise log lines with, documented in
        \sectionref{regex components}.

@@

@@

    \item [CLEANUP\_PROCESSING] \daemon{cleanup} processes every mail that
        passes through Postfix; details of what it does are available in
        \sectionref{Postfix Daemons}.  This action saves all data captured
        by the rule's regex if the log line has not come after a timeout
        (see \sectionref{discarding cleanup log lines}); it also creates
        the mail if necessary, setting its unknown origin flag (see
        \sectionref{pickup logging after cleanup}).

@@

    \item [CONNECT] Handle a remote client connecting: create a new
        connection, indexed by \daemon{smtpd} \acronym{pid}.  If a
        connection already exists it is treated as a symptom of a bug in
        \parsername{}, and the action will issue a warning containing the
        full contents of the existing connection plus the log line that has
        just been parsed.

@@

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check as
        described in \sectionref{out of order log lines}; this action, with
        cooperation from the COMMIT and SAVE\_DATA actions, deals with that
        complication.

@@

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  A new mail will be
        created, although out of order log entries may have caused the
        state table entry to already exist, as documented in
        \sectionref{pickup logging after cleanup}.

    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
        it kills all \daemon{smtpd} processes, requiring any active
        connections to be cleaned up, entered in the database, and deleted
        from the state tables.  Postfix probably kills all the other
        components too, but \parsername{} is only affected by \daemon{smtpd}
        processes exiting.

@@

    \item [SMTPD\_KILLED] Sometimes a \daemon{smtpd} is killed by a signal;
        this is processed in the same way as a \daemon{smtpd} dying, see
        the \texttt{SMTPD\_DIED} action for details.

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations; after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts that should occur in
        the intervening hours, e.g.\ timeouts for DNS requests or timeouts
        reading data from the client.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [TIMEOUT] The connection with the remote client timed out so the
        mail currently being transferred must be discarded.  The mail may
        have been accepted, in which case there's a data structure to
        dispose of, or it may not in which case there is not.  See
        \sectionref{timeouts during data phase} for full details.

@@

    \item [UNINTERESTING] This action just returns successfully: it is used
        when a log line needs to be parsed for completeness, but does not
        either provide any useful data to be saved or require anything to
        be done.

@@

@@

The role of the rules in the architecture is covered in detail in
\sectionref{rules in architecture}; this section is concerned with the
practical aspects of how rules are implemented and used in \parsername{}.
The structure of the rules table has already been documented in
\sectionref{rules table}; that abstract description will not be repeated
here, but should be fleshed out by the concrete example rule in
\sectionref{example rule in implementation}.  The process of creating new
rules from unparsed log lines is dealt with in \sectionref{creating new
rules in implementation}; the implementation of the utility supplied with
\parsername{} to create regexes from unparsed log lines is also described
in that section.  The rule conditions used in \parsername{} are explained
next, and the final portion of this section provides some
suggestions for how to detect overlapping rules.

@@

\label{regex components}

Each rule's regex will have keywords expanded when the rules are loaded,
for several reasons:

    \item It simplifies both reading and writing of regexes and makes each
        regex largely self-documenting.  E.g.\ the meaning of
        \_\_CLIENT\_HOSTNAME\_\_ is immediately clear, whereas
        \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered.

    \item It avoids needless repetition of complex regex components, and
        allows the components to be corrected or improved in one
        location.  E.g.\ \_\_SENDER\_\_ is used in 68 rules; if a mistake
        is discovered it can be corrected in one place only.

    \item It enables automatic extraction and saving of captured data.  The
        regex snippets use Perl 5.10's named capture buffers~\cite{perlre},
        so the mapping between captures and fields does not need to be
        explicitly specified by the rule.

For efficiency the keywords are expanded and every rule's regex is compiled
before attempting to parse the log file, otherwise every regex would be
recompiled each time it was used, resulting in a large, data dependent
slowdown, as shown in \sectionref{Caching compiled regexes}.  Most of the
keywords are named after the fields in the connections or results tables
they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_, \_\_DELAY\_\_,
\_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_, \_\_HELO\_\_,
\_\_MESSAGE\_ID\_\_, \_\_QUEUEID\_\_, \_\_RECIPIENT\_\_, \_\_SENDER\_\_,
\_\_SERVER\_HOSTNAME\_\_, \_\_SERVER\_IP\_\_, \_\_SIZE\_\_, and
\_\_SMTP\_CODE\_\_.

Some of the keywords need more explanation:

\begin{eqlist}

    \item [\_\_CHILD\_\_]  The queueid of a child mail; see
        \sectionref{Re-injected mails}.

    \item [\_\_COMMAND\_\_]  All \acronym{SMTP} commands, except
        \texttt{DATA}, which typically has a more specific rule.
        Priorities could have been used instead of excluding \texttt{DATA}.

    \item [\_\_CONN\_USE\_\_]  How many times the connection was reused;
        Postfix tries to reuse connections whenever possible to reduce the
        load on both the sending and receiving servers.

    \item [\_\_DATA\_\_]  This snippet is special: it matches a zero-length
        string, so it needs to be followed by a pattern that matches
        appropriately, but it captures whatever is matched so that it can
        be saved.  E.g.\ \verb!/(__DATA__connection (?:refused|reset!
        \newline{} \tab{}\verb!by peer))/! matches either ``connection
        refused'' or ``connection reset by peer'' and causes it to be saved
        to the data field of that log line's result.

    \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
        dies or is killed; see the SMTPD\_DIED action in
        \sectionref{actions in detail in implementation}.

    \item [\_\_RESTRICTION\_START\_\_]  Matches the start of a delivery
        rejected log line: XXX SHOULD I INCLUDE THIS\@? \newline{}
        \verb!/(__QUEUEID__): reject(?:_warning)?: ! \newline{}
        \verb!(?:RCPT|DATA) from ! \newline{}
        \verb!(?>(__CLIENT_HOSTNAME__)\\[)! \newline{}
        \verb!(?>(__CLIENT_IP__)\\]): (__SMTP_CODE__)! \newline{}
        \verb!(?: __ENHANCED_STATUS_CODE__)?/!

    \item [\_\_SHORT\_CMD\_\_]  Postfix sometimes logs \acronym{SMTP}
        commands in a short, single word form.

\end{eqlist}
